Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-12-20 manuscript.tex
latest

As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for this purpose. The lottery model gets its name from the way in which it represents competition: the growth of each type is determined by the number of contestants that it produces relative to the other types in the population (akin to a lottery),  weighted by a coefficient for each type representing its competitive ability. This is much simpler than having coefficients for the pairwise effects of types on each other (e.g. the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}. The classic lottery model  breaks down at low larval densities (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}. Here we analytically extend the classic lottery model to correctly account for low density behavior.

@@

Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories (no dispersal limitation). We assume adults cannot be ousted from occupied territories, so only propagules landing on occupied territories are included in $m_i$. More generally, $m_i$ only includes propagules which actually end up contesting unnocupied territories. For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.

@@

When multiple propagules land on the same territory, they compete to secure the territory as they develop. This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next iteration is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability. This includes cases where territories are won via a sexual contest.

>>>>

As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for this purpose. The lottery model gets its name from the way in which it represents competition: the growth of each type depends on how many territories it secures from a fixed territorial stock, which is determined by the number of contestants that it produces relative to the other types in the population weighted by a coefficient for each type representing competitive ability. This is much simpler than having coefficients for the pairwise effects of types on each other (e.g. the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}. The classic lottery model  breaks down at low absolute abundances at the scale of territorial contests (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}. Here we analytically extend the classic lottery model to correctly account for low density behavior.

@@

Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules''). We assume adults cannot be ousted from occupied territories, so only propagules landing on unoccupied territories are included in $m_i$. Dispersal is assumed to be unrestricted, so that all propagules have the same probability of landing on any of the $U$ unnocupied territories. Propagules disperse independently, and

 More generally, $m_i$ only includes propagules which actually end up contesting unnocupied territories. For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.

@@

When multiple propagules land on the same territory, they compete to secure the territory as they develop. This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a territory by the next iteration, assuming that at least one of its propagules is present, is $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability.
